---
title: "Mountain Pine Beetle Climatic Suitability Maps"
author: "Alex M. Chubaty, Barry J. Cooke, Eliot J. B. McIntire"
date: "October 26, 2016"
output: 
  pdf_document: 
    fig_caption: yes
    toc: yes
    toc_depth: 3
bibliography: climate-projections.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, cache.path = "cache/", echo = FALSE,
                      message = FALSE, warning = FALSE)

library(magrittr)
library(raster)
library(RCurl)

maps.dir <- if (Sys.info()[["sysname"]] == "Darwin") {
    '~/Documents/Data/shared/MPB/climate'
  } else if (Sys.info()[["sysname"]] == "Linux") {
    '~/Documents/Data/shared/MPB/climate'
  } else if (Sys.info()[["sysname"]] == "Windows") {
    '//W-VIC-A105388/shared/data/MPB/climate'
  }
if (!file.exists(maps.dir)) stop("maps dir does not exist.")

download = FALSE
```

## Overview

We have 7 MPB climate scenarios (using four different indices: `S`, `L`, `R`, `G`) ready to use as model drivers.
The Logan suitability index (`L`) is based on summer temperatures [@Logan:2003fr].
The Regniere suitability index (`R`) is based on MPB cold tolerance (*i.e.*, winter survival) [@Regniere:2007ip].
The Safranyik suitability index (`S`) is based on aspects of both summer temperatures and winter survival [@Safranyik:1975bk].
Finally, the composite SLR index (`G`) takes the geometric mean of the `S`, `L`, and `R` models.

These are described in further detail in @Nealis:2008gc (and the updated @Nealis:2014re?), and of course in their respective publications cited above.
BioSim was used to generate the maps [see @Bentz:2010bs; @Logan:2003fr; @Safranyik:2010ce].

```{r scenario-table}
scenarios <- data.frame(
  index = c('S', 'L', 'R', 'G'),
  description = c('Safranyik', 'Logan', 'Regniere', 'Composite'),
  stringsAsFactors = FALSE
)
knitr::kable(scenarios)
```

For each of the four indices, there are 1981-2010 normals plus the projections from two RCP scenarios (either 4.5 or 8.5), covering a span of 120 years, split into 30-year frames.

All maps are projected using LCC and cover all of Canada.

## Data source

Maps were produced in BioSim [see @Bentz:2010bs; @Logan:2003fr; @Safranyik:2010ce].

They were uploaded to and retrieved from the [NoFC FTP site](ftp://ftp.nofc.cfs.nrcan.gc.ca/uploads/MPB) on 26 October 2016.

```{r download-data, cache=FALSE}
if (download) {
  ## Get files of size zero
  file.size <- function(file) {
    return(file.info(file)$size)
  }
  
  eol <- ifelse(Sys.info()[["sysname"]] == "Windows", "\r\n", "\n")
  
  ## Define the ftp address (NOTE this address is temporary)
  ftp.site <- 'ftp://ftp.nofc.cfs.nrcan.gc.ca/uploads/MPB/'
  
  ftp.files <- getURL(ftp.site, ftp.use.epsv = FALSE, dirlistonly = TRUE) %>%
    strsplit(split = eol) %>% 
    unlist()
  
  for (f in ftp.files) {
    f.local <- file.path(maps.dir, f)
    f.ftp <- file.path(ftp.site, f)
    
    if (!file.exists(f.local)) {
      try(download.file(f.ftp, f.local))
    } else {
      if (file.size(f.local) == 0) {
        try(download.file(f.ftp, f.local))
      }
    }
  }
}
```

## Load maps

```{r load-maps, echo=TRUE}
files <- dir(path = maps.dir, pattern = '[.]tif', full.names = TRUE)

maps <- lapply(files, function(f) raster(f))
```

## Plot maps

```{r plot-maps}
#lapply(maps, plot)
```

# References

<!-- automatically generated using climate-projections.bib -->
